---
title: "week 1 exercises"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 1. R Markdown file

Create an R markdown file that, when knitted to an html documents, gives the same output as the file `R_markdown_exercise.html` (If you have Microsoft Word, check also that the output when knitted to Word agrees with `R_markdown_exercise.docx`)

NOTE: for submission, include your R Markdown file for this exercise as a separate file.

## Exercise 2. Rows and column operations

In this week's exercise you will make heavy use of the Palmer penguins dataset. A direct link to the dataset is <https://raw.githubusercontent.com/allisonhorst/palmerpenguins/refs/heads/main/inst/extdata/penguins.csv>

1. Load the Palmer penguin data as a data frame called `penguins`

2. Select only the penguins that resides in Biscoe Island. Assign the result to a new data frame called `Biscoe`, show a random sample of entries from this data frame.

3. How many penguins are recorded on the Biscoe island?

4. Create a new data frame called `penguins2`, wherein a new column called `bill_ratio` is defined as bill depth divided by bill length, and where the `sex` column is dropped. Display the first few entries of `penguins2`

5. What species of penguins appear in the database?

## Exercise 3. Dealing with missing data

Again load the Palmer penguins dataset as the data frame `penguins`

1. How many rows in the `penguins` dataset contains at least one missing value? (Note: missing values are coded as "NA" in the csv file, which is automatically converted into `NA` by `read_csv()`)

2. Recode all `NA` values in the `sex` column as `"unknown"`, call the result `penguins_re`. Display all columns of `penguins_re` for which sex is `"unknown"`

## Exercise 4: tidying data

We'll work with a subset of data from the World Health Organization Global Tuberculosis Report, which is included in `tidyr` under the variable name `who2`

```{r}
data(who2, package="tidyr")
glimpse(who2)
```

Besides "country" and "year", the columns represent different diagnostic code, and the cell corresponds to the count of people who are diagnosed. The diagnostic code consists of 3 separate section separated by underscore `_`:

 + the method of diagnosis (`rel` = relapse, `sn` = negative pulmonary smear, `sp` = positive pulmonary smear, `ep` = extrapulmonary),

 + gender (`f` = female, `m` = male), and

 + age group (`014` = 0-14 yrs of age, `1524` = 15-24, `2534` = 25-34, `3544` = 35-44 years of age, `4554` = 45-54, `5564` = 55-64, `65` = 65 years or older).

1. Convert the wide table into long table format. Call the column of diagnostic code "code" and the column of value "count". Call the result `who2_long`. 

2. Split the code into three columns "diagnosis", "gender", and "age_group". Call the result who2_tidy

## (Optional) Feedback questions

1. Are there any topic(s) from this week that still confuse you?

2. Which activities work and which doesn't? What do you want to see change and what do you want to keep?


