---
title: "R Dataframe Demo"
author: "Wing-Ho Ko"
output:
  html_notebook: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading packages for use

We'll load `readr` for importing external data as data frames and `dplyr` for data frame manipulation

When a package is loaded using the `library()` function, the functions it defines become available in the workspace and can be called directly. 

```{r}
library(readr)
library(dplyr)
```

## Loading external data into a dataframe

As a demonstration, we'll load the storms dataset (included in the dplyr package) in both online and offline manner.

To load the file directly from the internet, we note that the raw csv file can be reached from the link <https://raw.githubusercontent.com/tidyverse/dplyr/refs/heads/main/data-raw/storms.csv>. Using this as the argument to `read_csv()`, we do:

```{r}
read_csv("https://raw.githubusercontent.com/tidyverse/dplyr/refs/heads/main/data-raw/storms.csv")
```

Alternatively, download a copy of the csv file from <https://github.com/tidyverse/dplyr/blob/main/data-raw/storms.csv> and place it in the same folder as this markdown file. After that, we can load the file via:

```{r}
read_csv("storms.csv")
```

As with other data structure, you can assign a dataframe to a variable for future use and manipulation. In our case, we may do:

```{r}
storms <- read_csv("storms.csv")
```

Finally, since this is a dataset from a package we installed, we can also obtain the data via:

```{r}
data(storms, package="dplyr")
```


## Inspecting a dataframe

Unlike csv or other pure text data format, each column of a data frame has a specific data type. To inspect that, use the `glimpse()` function from `dplyr`, which also display useful information such as the number of rows and columns, as well as a preview of the data

```{r}
glimpse(storms)
```

In addition, if you just want to sample some rows from the data frame, you can use `head()`, `tail()`, and (with the `car` package) `some()`

(Note that instead of doing `library(car)`, we simply specify that the `some()` function should be loaded from the `car` package using `car::some()`)

```{r}
head(storms)
```

```{r}
tail(storms)
```

```{r}
car::some(storms)
```

To get the number of rows and columns programmatically (i.e., assignable to variables and thus can be further manipulated), use `nrow()` and `ncol()`

```{r}
nr <- nrow(storms)
nc <- ncol(storms)

print(c(nr, nc))
```

Extract the pressure values as a vector from the data frame

```{r}
pressure <- storms$pressure
pressure
```
