---
title: "R Dataframe Operations Demo"
author: "Wing-Ho Ko"
output:
  html_notebook: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading packages for use

This time we'll load the entire `tidyverse` meta package in one go. We will also load `janitor` to help clean up column names

```{r}
library(tidyverse)
library(janitor)
```


## Loading the dataset

Loading the storms dataset provided by dplyr

```{r}
data(storms, package="dplyr")
```

Inspect the dataset:

```{r}
glimpse(storms)
```


## Column operations

### `select()`

Extract only a few columns from the above data frame, as store the result in `storms_sub`

Note that data frames are "immutable". This means that you should treat `storms` and `storms_sub` as completely independent of each other.

```{r}
storms_sub <- select(storms, 
  c("name", "year", "month", "day", "hour", "wind", "pressure")
)
head(storms_sub)
```

Alternative syntax (often more convenient). Notice that column names are being treated like variable in this form

```{r}
storms_sub <- select(storms, name:hour | wind:pressure)
head(storms_sub)
```

### `mutate()`

Define a new column using `mutate()`

```{r}
storms_new <- mutate(storms_sub, wind_mph = wind * 1.151)
some(storms_new)
```

If the named argument coincide with the name of an existing column, the existing column is being overwritten instead

```{r}
storms_new <- mutate(storms_new, year = as.character(year))
car::some(storms_new)
```

### `rename()` and `relocate()`

Rename a column:

```{r}
storms_new <- rename(storms_new, wind_knots=wind)
car::some(storms_new)
```

reorder columns using relocate:

```{r}
storms_new <- relocate(storms_new, wind_mph, .after = wind_knots)
car::some(storms_new)
```

Perform the above transformations on our data frame in one go, and using the pipe operator `|>`

```{r}
storms_chained <- storms |> select(name:hour | wind:pressure) |> 
  mutate(wind_mph = wind * 1.151) |>
  mutate(year = as.character(year)) |>
  rename(wind_knots=wind) |>
  relocate(wind_mph, .after = wind_knots)

car::some(storms_chained)
```

## Row operations

### `filter()`

Filter rows of a data frame

```{r}
storms_sub |> filter(wind > 150)
```

Filter rows of a data frame using compound conditions

```{r}
storms_sub |> filter(wind > 150 & pressure < 900)
```

### `arrange()`

Arrange the rows of a data frame by the value of a single column

```{r}
storms_sub |> arrange(wind)
```

Arrange the rows of a data frame by a cascading set of columns

```{r}
storms_sub |> arrange(wind, pressure)
```

Display the storms with strongest wind by sorting wind in _descending_ order

```{r}
storms_sub |> arrange(desc(wind)) |> head()
```

### `distinct()`

Extract unique combination of storm name and year

```{r}
storms_sub |> distinct(name, year)
```

Extract the first appearance of each storm

```{r}
storms_sub |> arrange(year, month, day, hour) |>
  distinct(name, year, .keep_all=TRUE)
```

## Saving dataframe to .csv file or .rds file

Create the data frame to be saved

```{r}
storms_first <- storms_sub |> arrange(year, month, day, hour) |>
  distinct(name, year, .keep_all=TRUE)
```

Export the data frame as a .csv file

```{r}
write_csv(storms_first, "storms_first.csv")
```

Export the data frame to R native .rds file

```{r}
write_rds(storms_first, "storms_first.rda")
```

Load the saved rds file

```{r}
storms_from_rds <- read_rds("storms_first.rda")
some(storms_from_rds)
```

## cleaning up "bad" column names

Read a file with bad column names

```{r}
bad_names <- read_csv("data/bad_col_names.csv", name_repair="minimal")
glimpse(bad_names)
```

Clean up column names using `clean_names()` from `janitor`

```{r}
repaired_names <- bad_names |> clean_names()
glimpse(repaired_names)
```
