---
title: "Tidy data demo"
author: "Wing-Ho Ko"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r}
library(tidyverse)
```

## data tidying

### Convert a wide table to a long and tidy table

Inspect the `billboard` dataframe that comes with `tidyr`, which is a wide table that encode billboard ranks in the columns wk1, ..., wk76

```{r}
data(billboard, package="tidyr")
billboard
```

Pivot the wide table into long table:

```{r}
pivot_longer(billboard, wk1:wk76, names_to = "week", values_to = "rank") 
```

A bit more processing: remove the prefix "wk" from the "week" values, and convert into integer

```{r}
pivot_longer(
  billboard, wk1:wk76, 
  names_to = "week", names_prefix = "wk",
  values_to = "rank"
) |> mutate(week = as.integer(week))
```

### Convert crammed column to tidy columns

Import a summarized occurrence table

```{r}
occurrences <- read_csv("data/occurrences.csv")
```

```{r}
head(occurrences)
```

Separate scientific names into genus and species

```{r}
separate_wider_delim(
  occurrences, scientificName, " ", names=c("genus", "species")
)
```

#### convert an entity-attribute-value table into wider tidy table

Load the `economics_long` table (which originates from `ggplot2`)

```{r}
data(economics_long, package="ggplot2")
```

```{r}
glimpse(economics_long)
```

Check the collection of measurements

```{r}
economics_long |> distinct(variable)
```

Pivot into a wider tidy table

```{r}
economics <- pivot_wider(economics_long, id_cols="date", names_from=variable, values_from=value)

head(economics)
```
