---
title: "Tidy data demo"
author: "Wing-Ho Ko"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r}
library(tidyverse)
```

## Dealing with missing values

### Encode missing values when loading data

Download the Karachi, Pakistan sea level data from <https://psmsl.org/data/obtaining/stations/204.php>. The data encode missing values as -99999 and is a semicolon-separated file

```{r}
sea_level <- read_delim(
  "https://psmsl.org/data/obtaining/rlr.annual.data/204.rlrdata", delim=";",
  col_names = c("year", "depth_mm", "missing_days", "quality_flag"),
  na=c("", "NA", "-99999")
) |> mutate(
  year = as.numeric(year), 
  depth_mm = as.numeric(depth_mm)
)

sea_level
```

### Handle missing data

Load a simplified version of the storms data

```{r}
data(storms, package="dplyr")

storms_sub <- storms |> 
  select(name:hour | category | tropicalstorm_force_diameter)
```

Check the number of NA's for each row using `summary()`:

```{r}
summary(storms_sub)
```

Remove any observations that have NA's in them

```{r}
storms_notna <- drop_na(storms_sub)
summary(storms_notna)
```

Count the number of rows removed, and note that it is bigger than the number of NA's in any individual columns

```{r}
nrow(storms_sub) - nrow(storms_notna)
```

Remove only the rows where category is NA:

```{r}
weather_cat = drop_na(storms_sub, category)
summary(weather_cat)
```

Construct a data frame consisting only of entries for which category **is missing**:

```{r}
storms_isna = storms_sub |> filter(is.na(category))
car::some(storms_isna)
```

### Replacing missing data

As a simple (but probably conceptually wrong) example, let's replace missing `depth_mm` data in the sea_level data set with mean value

```{r}
mean_sea_level <- mean(sea_level$depth_mm, na.rm=TRUE)
mean_sea_level
```

```{r}
sea_level <- sea_level |> 
  mutate(depth_mm = replace_na(depth_mm, mean_sea_level))

sea_level
```


### Complete implicit data

Load the Ballard Locks salmon count data

```{r}
salmons <- read_csv("data/Ballard_salmon_counts.csv")
```

```{r}
salmons
```

Fill in missing months

```{r}
salmons_filled <- salmons |> complete(year, month=seq(1, 12)) 
```

```{r}
salmons_filled
```
